Algorithmic Approaches to Match Degraded Land Impressions

### 2017-10-03 degraded-paper.tex
Paper updates

\setlength{\droptitle}{-2em}
  \title{Algorithmic Approaches to Match Degraded Land Impressions}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{Eric Hare, Heike Hofmann, Alicia Carriquiry}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{9/5/2017}

@@

Intense scrutiny has been focused on the process of bullet matching in
recent years (e.g., Giannelli 2011). Bullet matching, the process of
determining whether two bullets could have been fired from the same gun
barrel, has traditionally been performed without meaningful
determination of error rates or statistical assessments of uncertainty
(National Research Council 2009). There have been some attempts towards
developing mathematical and statistical approaches to bullet
matching.One such attempt was the definition of CMS, the Consecutively
Matching Striae (Biasotti 1959), with a cutoff of six to separate
matches from non-matches. Still, rigorous assessments of the
applicability of such cutoffs have not to this point been described
(Advisors on Science and Technology 2016).

Recently, several authors have addressed these well-known shortcomings.
Focusing on firing pin impressions and breech faces, Riva and Champod
(2014) have described an automated algorithm using 3D images that
enables comparison between pairs of exemplars. Other examples of work in
this and related areas include Petraco and Chan (2012), W. Chu et al.
(2011), T. Vorburger et al. (2011), and others. In our approach to this
problem, Automatic Matching of Bullet Lands, we used the Hamby 252 set
(Hamby, Brundage, and Thorpe 2009) to train and develop a random forest
in order to provide a matching probability for two bullet lands (Hare,
Hofmann, and Carriquiry 2016). While the algorithm had a very strong
performance on this set, some limitations were immediately clear. For
instance, performance was assessed only on this single set of 35 bullets
fired from a consecutively manufactured set of only ten gun barrels.
Each of these bullets was part of controlled study, and the full lands
were available for matching. While there were some data quality issues,
this was still a near ideal test case for the algorithm.

Real world applications of bullet matching often involve the recovery of
fragments of bullets from the crime scene. Traditional features used in
forensic examination work well for a full land, but there has been less
investigation into their performance in the case of a fragmented land.
For example, the CMS is naturally limited by the portion of the land
that can be recovered.

@@

To start, we introduce a standardized version of each of the features
used in the matching routine proposed by (Hare, Hofmann, and Carriquiry
2016). These features are computed on \emph{aligned pairs of bullet land
impressions} rather than on individual lands. This enables us, for
instance, to compute the number of matching striae. Table \ref{tab:ccf1}
and Table \ref{tab:ccf2} provide an example of six land-to-land
comparisons and the derived features for these comparisons. The two
\texttt{profile\_id} columns identify a particular land from the two
Hamby datasets, and the remaining columns are the derived features for
these comparisons.

\begin{table}[ht]
\centering
\begin{tabular}{rrrrrrr}
  \hline
profile1\_id & profile2\_id & ccf & rough\_cor & lag & D & sd\_D \\ 
  \hline
49 & 540 & 0.2689 & 0.1091 & -0.1734 & 0.0026 & 0.0043 \\ 
  49 & 1032 & 0.3878 & 0.3582 & 0.1938 & 0.0017 & 0.0030 \\ 
  49 & 1540 & 0.2410 & 0.1320 & 0.2781 & 0.0023 & 0.0039 \\ 
  49 & 2044 & 0.2892 & 0.1136 & 0.0828 & 0.0023 & 0.0038 \\ 
  49 & 2583 & 0.1769 & 0.0214 & -0.0563 & 0.0026 & 0.0041 \\ 
  49 & 3074 & 0.2715 & 0.1254 & 0.0938 & 0.0020 & 0.0033 \\ 
   \hline
\end{tabular}
\caption{A sample of six land-to-land comparisons, and derived features for these comparisons.} 
\label{tab:ccf1}
\end{table}

\begin{table}[ht]
\centering
\begin{tabular}{rrrrrrr}
  \hline
signature\_length & overlap & matches & mismatches & cms & non\_cms & sum\_peaks \\ 
  \hline
1.9328 & 0.9103 & 1.1368 & 10.2251 & 0.5684 & 3.7182 & 2.0038 \\ 
  1.9359 & 0.9193 & 5.0571 & 6.5593 & 3.3714 & 2.3426 & 6.5264 \\ 
  1.9359 & 0.8757 & 1.7696 & 10.3662 & 1.1797 & 5.8592 & 2.0192 \\ 
  1.9172 & 0.9829 & 1.5920 & 7.9441 & 0.5307 & 3.4756 & 1.8139 \\ 
  1.8234 & 0.9692 & 1.1317 & 8.8479 & 0.5659 & 4.9155 & 0.8555 \\ 
  1.9172 & 0.9690 & 2.6913 & 10.7645 & 1.0765 & 3.4251 & 3.5272 \\ 
   \hline
\end{tabular}
\caption{The remaining derived features for the previous six land-to-land comparisons.} 
\label{tab:ccf2}
\end{table}

We generalize the definitions of these features to account for the
possibility that we may be handling degraded bullet lands, where only
fragments can be recovered. The definition of each feature is given
below, where \(f(t)\) represents the height values of the first profile,
and \(g(t)\) the height values of the second:

\begin{itemize}
\tightlist
\item
  \textbf{ccf} (\%) is the maximum value of the Cross-Correlation
  function evaluated at the optimal alignment. The Cross-Correlation
  function is defined as
  \(C(\tau) = \int_{-\infty}^{\infty} f(t)g(t + \tau)dt\) where \(\tau\)
  represents the the lag of the second signature (T. Vorburger et al.
  2011).
\item
  \textbf{rough\_cor} (\%) is a new feature that quantifies the
  correlation between the two signatures after performing a second LOESS
  smoothing stage and then subtracting the result from the original
  signatures. This attempts to model the roughness of the surface after
  removing structure such as waviness.
\item
  \textbf{lag} (mm) Is the optimal lag for the ccf value.
\item
  \textbf{D} (mm) is the Euclidean vertical distance between each height
  value of the aligned signatures. This is defined as
  \(D^2 = \frac{1}{\text{\#}t}\sum_t \left[f(t) - g(t)\right]^2\). This
  is a measure of the total variation between two functions (Clarkson
  and Adams 1933).
\item
  \textbf{sd\_D} (mm) provides the standard deviation of the values of
  \emph{D} from above.
\item
  \textbf{signature\_length} (mm) is the overall length of the smallest
  of the two aligned signatures.
\item
  \textbf{overlap} (\%) provides the percentage of the two signatures
  that overlap after the alignment stage.
\item
  \textbf{matches} (per mm) is the number of matching peaks/valleys
  (striae) per millimeter of the overlapping portion of the aligned
  signatures.
\item
  \textbf{mismatches} (per mm) is the number of mismatching
  peaks/valleys (striae) per millimeter of the overlapping portion of
  the aligned signatures.
\item
  \textbf{cms} (per mm) is the number of consecutively matching
  peaks/valleys (striae) per millimeter of the overlapping portion of
  the aligned signatures (Biasotti 1959, Wei Chu et al. (2013)).
\item
  \textbf{non\_cms} (per mm) is the number of consecutive mismatching
  peaks/valleys (striae) per millimeter of the overlapping portion of
  the aligned signatures.
\item
  \textbf{sum\_peaks} (per mm) is the the sum of the average heights of
  matched striae.
\end{itemize}

@@

We can observe the distributions of both CCF and the Roughness
Correlation side by side, differentiating between known matches and
known non-matches. Figure \ref{fig:roughcordist} shows this. It can be
seen that the separation of the two groups along both variables is quite
strong, but many known-matches have low values for each. The
distributions of known non-matches, on the other hand, is relatively
symmetric and normal.

\begin{figure}[htbp]
\centering
\includegraphics{degraded-paper_files/figure-latex/roughcordist-1.pdf}
\caption{\label{fig:roughcordist}Distributions of the Roughness Correlation
compared to the CCF for known matches and known non-matches}
\end{figure}

@@

Using these features, we can train a randomForest (Liaw and Wiener 2002)
model which attempts to predict whether two lands match given the value
of the features. There are currently three studies for bullets included
in the NIST ballistics database (REF). Those are Hamby (Set 252), Hamby
(Set 44), and Cary. For purposes of analysis we describe in this paper,
we exclude the Cary bullets from consideration, because the study was
designed to assess the persistence of striation markings over a series
of fires from the same barrel. Thus, every Cary bullet is a known match
to every other Cary bullet, although early firings do not produce the
same markings that later firings do. Hence, we will consider Hamby (Set
252) and Hamby (Set 44) only.

We can now train the forest using the features we defined earlier. We
split the data into an 80\% training, 20\% testing sets to assess the
algorithm's out of sample performance, using the \texttt{caret} package
(Jed Wing et al. 2016). Table \ref{tab:avgforest} displays the results
in the form of a confusion matrix on the test set, averaged over ten
independent random forests trained on ten random data subsets. There are
a total of 83,028 land-to-land comparisons, of which 1208 are among
known matching land impressions and 81,820 are among known non-matching
land impressions. It can be seen that false positives are exceedingly
rare, but false negatives occur more frequently (approximately 65 false
negative land to land comparisons on the test set, compared with an
average of less than four false positives).

\begin{table}[H]
\centering
\begin{tabular}{lr}
  \hline
Result & Count \\ 
  \hline
False Negative & 65.3 \\ 
  False Positive & 3.8 \\ 
  True Negative & 16363.1 \\ 
  True Positive & 171.8 \\ 
   \hline
\end{tabular}
\caption{The average confusion matrix for the 10 random forests. It can be seen that false positives are exceedingly rare, but false negatives occur more frequently.} 
\label{tab:avgforest}
\end{table}

These results suggest that our algorithm is too conservative in
predicting a match when in fact the bullets were fired from the same gun
barrel. We can break down the confusion matrix by the study from which
each of the two land impressions originated. Table
\ref{tab:avgforeststudy} shows the average confusion matrix for the 10
random forests, broken down by study. It can be seen that Hamby252 to
Hamby252 comparisons exhibit the fewest errors, while Hamby252 to
Hamby44 comparisons exhibit the most errors on average. This intuitively
makes some sense given the potential presence of scanner operator
effects, which we address further in this section.

\begin{table}[H]
\centering
\begin{tabular}{lllll}
  \hline
Study & False Negative & False Positive & True Negative & True Positive \\ 
  \hline
Hamby252\_Hamby252 & 0.29\% & 0.01\% & 98.77\% & 0.93\% \\ 
  Hamby252\_Hamby44 & 0.47\% & 0.02\% & 98.35\% & 1.16\% \\ 
  Hamby44\_Hamby44 & 0.34\% & 0.04\% & 98.72\% & 0.9\% \\ 
   \hline
\end{tabular}
\caption{The average confusion matrix for the 10 random forests, broken down by study. It can be seen that Hamby252 to Hamby252 comparisons exhibit the fewest errors, while Hamby252 to Hamby44 comparisons exhibit the most on average.} 
\label{tab:avgforeststudy}
\end{table}

@@

As a first stage to assessing feature robustness, we produce parallel
coordinate plots of the various features based on true positives, true
negatives, false positives, and false negatives. Figures \ref{fig:pcp}
and \ref{fig:pcp2} display these plots. The means of the true positive
and the false negative groups are shown, respectively, in the two
figures. False positives tend to occur with anomalously high values of
\texttt{sum\_peaks}, \texttt{matches}, and \texttt{ccf}, while false
negatives tend to resemble very closely the feature distribution of the
true negatives, with a slightly higher average \texttt{ccf}.

\begin{figure}[htbp]
\centering
\includegraphics{degraded-paper_files/figure-latex/pcp-1.pdf}
\caption{\label{fig:pcp}Parallel coordinate plot of the features based on
the random forest confusion matrix for true and false positives.}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{degraded-paper_files/figure-latex/pcp2-1.pdf}
\caption{\label{fig:pcp2}Parallel coordinate plot of the features based on
the random forest confusion matrix for true and false negatives.}
\end{figure}

@@

We attempt to quantify the effect of the study on the matching
probability by fitting a new random forest which is designed to predict
the study based on the derived features. Ideally, if the assumption of
independence between lands holds across different operators, this forest
should have poor performance - The set of derived features should be
relatively consistent among known matches and known non-matches
regardless of the study since the Hamby data in both sets originated
from the same gun barrels.

Table \ref{tab:studypred} shows the confusion matrix, with column
proportions, for the random forest with study as the response. It can be
seen that while overall the random forest performs poorly, as hoped,
comparisons between Hamby252 bullets are more distinguishable from other
comparisons.

\begin{table}[H]
\centering
\begin{tabular}{llll}
  \hline
Prediction $\backslash$ Actual & Hamby252\_Hamby252 & Hamby252\_Hamby44 & Hamby44\_Hamby44 \\ 
  \hline
Hamby252\_Hamby252 & 28.5\% & 10.23\% & 3.24\% \\ 
  Hamby252\_Hamby44 & 70.57\% & 83.27\% & 79.78\% \\ 
  Hamby44\_Hamby44 & 0.93\% & 6.51\% & 16.98\% \\ 
   \hline
\end{tabular}
\caption{Confusion Matrix (Column Proportions) for the random forest with study as the response. It can be seen that while overall the random forest performs poorly, as hoped, comparisons between Hamby252 bullets is more distinguishable from other comparisons.} 
\label{tab:studypred}
\end{table}

Figure \ref{fig:ccfstudy} shows the distributions of the features
defined above, faceted by whether the lands are known to be fired from
the same gun barrel, across different study to study comparisons. The
distributions among the known non-matches seem relatively consistent
across study based on visual inspection. On the other hand, among known
matches, Hamby252 to Hamby252 comparisons exhibit more pronounced
features, including a higher average ccf, higher number of matches, and
higher value of sum\_peaks.

@@

Though visual inspection clearly shows differences, we can more formally
assess the differences between distributions with a Kolmogrov-Smirnov
test. Table \ref{tab:kstests} gives the results of pairwise tests, for
each feature, between different set comparisons, and between known
matches compared with known non-matches. Although the tests are
significant, looking at the raw values of the D statistic suggest that
the largest effect sizes do in fact occur in comparisons with two
Hamby252 lands, as the visual inspection of the boxplots also suggested.

@@

Indeed, another Kolmogorov-Smirnov test confirms a significant
difference in the distributions of these values
(\(D = 0.6239, p < 0.0001\)). This result strongly suggests that the
operator effect in the bullet scanning procedure must be taken into
account in order to assume pairwise independence.

@@

Figure \ref{fig:featexp} gives feature expression for known matches, as
a function of the proportion of land impression recovered. It is
immediately obvious that the variability in feature expression is large
when only a small fraction of the land is recovered, such as 25\%. For
instance, \texttt{sum\_peaks} and \texttt{cms} both drop, while
\texttt{D} rises. Interestingly, some of the features are better
expressed for the middle-fixed case. Overall, feature expression remains
relatively consistent as long as we recover 50\% or more of the land
impression. Feature \ref{fig:featexp2} shows the feature expression for
known non-matches by comparison.

@@

\begin{figure}[htbp]
\centering
\includegraphics{degraded-paper_files/figure-latex/featexp2-1.pdf}
\caption{\label{fig:featexp2}Feature expression for known non-matches, as a
function of land proportion. Patterns of feature expression as a
function are far less visible in this case.}
\end{figure}

To come full circle, we now attempt to match a particular land which
exhibits bad tank rash. Figure \ref{fig:br924} provides an image of the
surface of this land impression. Due to the tank rash, this particular
land impression was originally excluded from consideration (see (Hare,
Hofmann, and Carriquiry 2016)). However, it appears that approximately
half of the bullet land remains relatively unaffected. We extract a
signature from the unaffected half and attempt to match this signature
to its full known match.

@@

\begin{align}
P(M) &= 1 - P(NM) \\
     &= 1 - (P(NM1) \times P(NM2) \times ... \times P(NM6)) \\
     &= 1 - ((1 - P(M1)) \times (1 - P(M2)) \times ... \times (1 - P(M6)))
\end{align}

>>>>

\setlength{\droptitle}{-2em}
  \title{Algorithmic Approaches to Match Degraded Land Impressions}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{Eric Hare, Heike Hofmann, Alicia Carriquiry}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{10/05/2017}

@@

Intense scrutiny has been focused on the process of bullet matching in
recent years (e.g., Giannelli 2011). Bullet matching, the process of
determining whether two bullets could have been fired from the same gun
barrel, has traditionally been performed without meaningful
determination of error rates or statistical assessments of uncertainty
(National Research Council 2009). There have been some attempts towards
developing mathematical and statistical approaches to bullet matching.
One such attempt was the definition of CMS, the Consecutively Matching
Striae (Biasotti 1959), with a cutoff of six to separate matches from
non-matches. Still, rigorous assessments of the applicability of such
cutoffs have not to this point been described (Advisors on Science and
Technology 2016).

Recently, several authors have addressed these well-known shortcomings.
Focusing on firing pin impressions and breech faces, Riva and Champod
(2014) have described an automated algorithm using 3D images that
enables comparison between pairs of exemplars. Other examples of work in
this and related areas include Petraco and Chan (2012), W. Chu et al.
(2011), T. Vorburger et al. (2011), and others. In our approach to this
problem, Automatic Matching of Bullet Lands, we used the Hamby 252 set
(Hamby, Brundage, and Thorpe 2009) to train and develop a random forest
in order to provide a matching probability for two bullet lands (Hare,
Hofmann, and Carriquiry 2016). While the algorithm had a very strong
performance on this set, some limitations were immediately clear. For
instance, performance was assessed only on this single set of 35 bullets
fired from a consecutively manufactured set of only ten known and 15
unknown gun barrels. Each of these bullets was part of controlled study,
and the full lands were available for matching. While there were some
data quality issues, this was still a near ideal test case for the
algorithm.

Real world applications of bullet matching often involve the recovery of
fragments of bullets from the crime scene (Council and others 2004).
Traditional features used in forensic examination work well for a full
land, but there has been less investigation into their performance in
the case of a fragmented land. For example, the CMS is naturally limited
by the portion of the land that can be recovered and varies across
manufacturers (W. Chu et al. 2011).

@@

To start, we introduce a standardized version of each of the features
used in the matching routine proposed by (Hare, Hofmann, and Carriquiry
2016). These features are computed on \emph{aligned pairs of bullet land
impressions} rather than on individual lands. This enables us, for
instance, to compute the number of matching striae between two lands. We
generalize the definitions of these features to account for the
possibility that we may be handling degraded bullet lands, where only
fragments can be recovered. The definition of each feature is given
below, where \(f(t)\) represents the height values of the first profile
at position \(t\) along the signature, and \(g(t)\) the height values of
the second. An indication of whether the feature is new since Hare,
Hofmann, and Carriquiry (2016) is also given:

\begin{itemize}
\tightlist
\item
  \textbf{ccf} (\%) is the maximum value of the Cross-Correlation
  function evaluated at the optimal alignment. The Cross-Correlation
  function is defined as
  \(C(\tau) = \int_{-\infty}^{\infty} f(t)g(t + \tau)dt\) where \(\tau\)
  represents the the lag of the second signature (T. Vorburger et al.
  2011).
\item
  \textbf{rough\_cor} (new) (\%) feature that quantifies the correlation
  between the two signatures after performing a second LOESS smoothing
  stage and then subtracting the result from the original signatures.
  This attempts to model the roughness of the surface after removing
  structure such as waviness.
\item
  \textbf{lag} (mm) Is the optimal lag for the ccf value.
\item
  \textbf{D} (mm) is the Euclidean vertical distance between each height
  value of the aligned signatures. This is defined as
  \(D^2 = \frac{1}{\text{\#}t}\sum_t \left[f(t) - g(t)\right]^2\). This
  is a measure of the total variation between two functions (Clarkson
  and Adams 1933).
\item
  \textbf{sd\_D} (mm) provides the standard deviation of the values of
  \emph{D} from above.
\item
  \textbf{signature\_length} (mm) is the overall length of the smallest
  of the two aligned signatures.
\item
  \textbf{overlap} (new) (\%) provides the percentage of the two
  signatures that overlap after the alignment stage.
\item
  \textbf{matches} (per mm) is the number of matching peaks/valleys
  (striae) per millimeter of the overlapping portion of the aligned
  signatures.
\item
  \textbf{mismatches} (per mm) is the number of mismatching
  peaks/valleys (striae) per millimeter of the overlapping portion of
  the aligned signatures.
\item
  \textbf{cms} (per mm) is the number of consecutively matching
  peaks/valleys (striae) per millimeter of the overlapping portion of
  the aligned signatures (Biasotti 1959, Wei Chu et al. (2013)).
\item
  \textbf{non\_cms} (per mm) is the number of consecutive mismatching
  peaks/valleys (striae) per millimeter of the overlapping portion of
  the aligned signatures.
\item
  \textbf{sum\_peaks} (per mm) is the the sum of the average heights of
  matched striae.
\end{itemize}

@@

We can observe the distributions of both CCF and the Roughness
Correlation side by side, differentiating between known matches and
known non-matches. Figure \ref{fig:roughcordist} displays this as an
empirical CDF plot. It can be seen that the separation of known matches
and known non-matches along both the CCF and the Roughness Correlation
is quite strong and follows similar distributions (the known non-matches
are relatively symmetric, while the known matches are very skewed left).
However, some known non-matches with CCF values that would typically be
indicative of a match have relatively lower values for the Roughness
Correlation, which indicates that this feature could provide some added
value when it comes to discriminating between matches and non-matches.

\begin{figure}[htbp]
\centering
\includegraphics{degraded-paper_files/figure-latex/roughcordist-1.pdf}
\caption{\label{fig:roughcordist}Empirical CDFs of the Roughness Correlation
compared to the CCF for known matches and known non-matches. It can be
seen that the distribution of each feature for the known non-matches is
quite symmetric, while the distribution for each feature for the known
matches is skewed left.}
\end{figure}

@@

Using these features, we can train a randomForest (Liaw and Wiener 2002)
model which attempts to predict whether two lands match given the value
of the features. There are currently three studies for bullets included
in the NIST Ballistics Toolmark Research Database. Those are Hamby (Set
252), Hamby (Set 44), and Cary. For purposes of the analysis we describe
in this paper, we exclude the Cary bullets from consideration, because
the study was designed to assess the persistence of striation markings
over a series of fires from the same barrel. Thus, every Cary bullet is
a known match to every other Cary bullet. Hence, we will consider Hamby
(Set 252) and Hamby (Set 44) only. This leaves us a total of 83,028
land-to-land comparisons, of which 1208 are among known matching land
impressions and 81,820 are among known non-matching land impressions.

We can now train the forest using the features we defined earlier. Using
the \texttt{caret} package (Jed Wing et al. 2016), we perform the
following partitioning scheme. Out of the 50 barrels total, ten knowns
and fifteen unknowns from each of the two Hamby sets, we hold out ten
barrels randomly as a testing set, and use the remaining 40 to train the
model. We repeat this procedure ten times and average the confusion
matrix in order to assess the model accuracy with different holdout
samples. Table \ref{tab:avgforest} displays the results in the form of a
confusion matrix on the test set, averaged over these ten independent
random forests trained on ten random barrel subsets. It can be seen that
false positives are exceedingly rare, but false negatives occur more
frequently (approximately 21` false negative land to land comparisons on
the test set, compared with an average of less than two false
positives).

\begin{table}[H]
\centering
\begin{tabular}{lr}
  \hline
Result & Count \\ 
  \hline
False Negative & 20.7 \\ 
  False Positive & 1.9 \\ 
  True Negative & 3716.8 \\ 
  True Positive & 56.5 \\ 
   \hline
\end{tabular}
\caption{The average confusion matrix for the 10 random forests. It can be seen that false positives are exceedingly rare, but false negatives occur more frequently.} 
\label{tab:avgforest}
\end{table}

These results suggest that our algorithm is too conservative in
predicting a match when in fact the bullets were fired from the same gun
barrel. We can break down the confusion matrix by the study from which
each of the two land impressions originated. Table
\ref{tab:avgforeststudy} shows the average confusion matrix for the 10
random forests, broken down by study. It can be seen that Hamby252 to
Hamby252 comparisons exhibit the fewest errors, while Hamby44 to Hamby44
comparisons exhibit the most errors on average. This intuitively makes
some sense given the potential presence of scanner operator effects,
which we address further in this section.

\begin{table}[H]
\centering
\begin{tabular}{lllll}
  \hline
Study & False Negative & False Positive & True Negative & True Positive \\ 
  \hline
Hamby252\_Hamby252 & 0.39\% & 0.02\% & 97.56\% & 2.03\% \\ 
  Hamby252\_Hamby44 & 0.45\% & 0.05\% & 98.31\% & 1.19\% \\ 
  Hamby44\_Hamby44 & 0.89\% & 0.09\% & 97.68\% & 1.34\% \\ 
   \hline
\end{tabular}
\caption{The average confusion matrix for the 10 random forests, broken down by study. It can be seen that Hamby252 to Hamby252 comparisons exhibit the fewest errors, while Hamby44 to Hamby44 comparisons exhibit the most on average.} 
\label{tab:avgforeststudy}
\end{table}

@@

Our goal is to assess the robustness of the previously defined features
as it pertains to our bullet matching routines. This goal is both a
backward looking assessment of our previous results for full
land-to-land comparisons, and a forward looking one to help support the
case that these can be used in the degraded land case. As a first stage
to assessing this robustness, we produce parallel coordinate plots of
the various features based on true positive, true negative, false
positive, and false negative land-to-land matches. Figure \ref{fig:pcp}
displays these plots. The means of the true positive and the true
negative groups are shown as thick blue lines, respectively, in the two
panels. The dashed lines represent individual land to land comparisons,
with errors highlighted larger in red. It can be seen that the few false
positives tendt o have anomalously high \texttt{ccf} or
\texttt{matches}, while the false negatives tend have a lot of
variability, though tending to also have a high \texttt{ccf} value.

\begin{figure}[htbp]
\centering
\includegraphics{degraded-paper_files/figure-latex/pcp-1.pdf}
\caption{\label{fig:pcp}Parallel coordinate plot of the features based on
the random forest confusion matrix for true and false positives (above),
and true and false negatives (below). False positives tend to have some
feature anomalously high, while false negatives exhibit quite a spread,
sometimes having very large values of CCF or CMS, for example.}
\end{figure}

@@

We attempt to quantify the effect of the study on the matching
probability by fitting a new random forest which is designed to predict
the study from which the scans came from based on the derived features.
It should be noted that the two sets of scans were performed by two
trained professionals at NIST, and therefore this analysis is intended
to shed light on how slightly different operating procedures for the
scans may lead to varying results in algorithms derived from these
scans. Ideally, if the assumption of independence between lands holds
across different operators, this forest should have poor performance -
The set of derived features should be relatively consistent among known
matches and known non-matches regardless of the study since the Hamby
data in both sets originated from the same gun barrels.

Table \ref{tab:studypred} shows the confusion matrix, with column
proportions, for the random forest with study as the response. It can be
seen that indeed the random forest performs poorly, as hoped, indicating
that a simple model to predict the study using the features available is
not enough to detect the operator effects.

\begin{table}[H]
\centering
\begin{tabular}{llll}
  \hline
Prediction $\backslash$ Actual & Hamby252\_Hamby252 & Hamby252\_Hamby44 & Hamby44\_Hamby44 \\ 
  \hline
Hamby252\_Hamby252 & 9.6\% & 8.32\% & 11.28\% \\ 
  Hamby252\_Hamby44 & 81.47\% & 81.06\% & 78.68\% \\ 
  Hamby44\_Hamby44 & 8.93\% & 10.63\% & 10.04\% \\ 
   \hline
\end{tabular}
\caption{Confusion Matrix (Column Proportions) for the random forest with study as the response. It can be seen that the random forest performs poorly, as hoped, indicating that a simple model to predict the study using the features available is not enough to detect the operator effects.} 
\label{tab:studypred}
\end{table}

Figure \ref{fig:ccfstudy} shows the distributions of the land-to-land
features, faceted by whether the lands are known to be fired from the
same gun barrel, across different study to study comparisons. The
distributions among the known non-matches seem relatively consistent
across study based on visual inspection. On the other hand, among known
matches, Hamby252 to Hamby252 comparisons exhibit more pronounced
features, including a higher average ccf, higher number of matches, and
higher value of sum\_peaks.

@@

Though visual inspection clearly shows differences, we can more formally
assess the differences between distributions with a Kolmogrov-Smirnov
test. Table \ref{tab:kstests} gives the results of pairwise tests, for
each feature, between different set comparisons, and between known
matches compared with known non-matches. Although most of the tests are
significant, looking at the raw values of the D statistic suggest that
the largest effect sizes do in fact occur in comparisons with two
Hamby252 lands, as the visual inspection of the boxplots also suggested.

\begin{table}[H]
\centering
\begin{tabular}{llllrlr}
  \hline
set1 & set2 & feature & matchtest & matchd & nonmatchtest & nonmatchd \\ 
  \hline
H252\_H252 & H252\_H44 & ccf & $<$ 0.0001 & 0.2723 & 1e-04 & 0.0189 \\ 
  H252\_H252 & H252\_H44 & cms & $<$ 0.0001 & 0.1751 & $<$ 0.0001 & 0.0245 \\ 
  H252\_H252 & H252\_H44 & D & $<$ 0.0001 & 0.2567 & $<$ 0.0001 & 0.1049 \\ 
  H252\_H252 & H252\_H44 & matches & $<$ 0.0001 & 0.1933 & $<$ 0.0001 & 0.0327 \\ 
  H252\_H252 & H252\_H44 & mismatches & $<$ 0.0001 & 0.2015 & 0.3537 & 0.0079 \\ 
  H252\_H252 & H252\_H44 & overlap & 0.0492 & 0.0984 & $<$ 0.0001 & 0.0276 \\ 
  H252\_H252 & H252\_H44 & rough\_cor & $<$ 0.0001 & 0.2647 & $<$ 0.0001 & 0.0970 \\ 
  H252\_H252 & H252\_H44 & sum\_peaks & 8e-04 & 0.1426 & 0.0015 & 0.0162 \\ 
  H252\_H252 & H44\_H44 & ccf & $<$ 0.0001 & 0.2160 & $<$ 0.0001 & 0.0257 \\ 
  H252\_H252 & H44\_H44 & cms & $<$ 0.0001 & 0.2515 & $<$ 0.0001 & 0.0467 \\ 
  H252\_H252 & H44\_H44 & D & $<$ 0.0001 & 0.2342 & $<$ 0.0001 & 0.1946 \\ 
  H252\_H252 & H44\_H44 & matches & $<$ 0.0001 & 0.2770 & $<$ 0.0001 & 0.0713 \\ 
  H252\_H252 & H44\_H44 & mismatches & $<$ 0.0001 & 0.2505 & 0.0414 & 0.0138 \\ 
  H252\_H252 & H44\_H44 & overlap & 0.2432 & 0.0906 & $<$ 0.0001 & 0.0408 \\ 
  H252\_H252 & H44\_H44 & rough\_cor & $<$ 0.0001 & 0.2242 & $<$ 0.0001 & 0.1718 \\ 
  H252\_H252 & H44\_H44 & sum\_peaks & 1e-04 & 0.1926 & $<$ 0.0001 & 0.0289 \\ 
  H252\_H44 & H44\_H44 & ccf & 0.1149 & 0.0883 & $<$ 0.0001 & 0.0259 \\ 
  H252\_H44 & H44\_H44 & cms & 0.111 & 0.0888 & $<$ 0.0001 & 0.0262 \\ 
  H252\_H44 & H44\_H44 & D & 0.2923 & 0.0724 & $<$ 0.0001 & 0.0906 \\ 
  H252\_H44 & H44\_H44 & matches & 0.0603 & 0.0977 & $<$ 0.0001 & 0.0423 \\ 
  H252\_H44 & H44\_H44 & mismatches & 0.3301 & 0.0700 & 0.1633 & 0.0096 \\ 
  H252\_H44 & H44\_H44 & overlap & 0.8231 & 0.0465 & 1e-04 & 0.0190 \\ 
  H252\_H44 & H44\_H44 & rough\_cor & 0.2671 & 0.0741 & $<$ 0.0001 & 0.0769 \\ 
  H252\_H44 & H44\_H44 & sum\_peaks & 0.047 & 0.1011 & 0.006 & 0.0147 \\ 
   \hline
\end{tabular}
\caption{Results for the Kolmogrov-Smirnov distributional test.} 
\label{tab:kstests}
\end{table}

@@

Indeed, another Kolmogorov-Smirnov test confirms a significant
difference in the distributions of these values
(\(D = 0.6239, p < 0.0001\)). This result strongly suggests that the
operator effect in the bullet scanning procedure must be taken into
account in order to assume pairwise independence of bullet land scans
between Hamby sets 252 and 44.

@@

Figure \ref{fig:featexp} gives feature expression for known matches, as
a function of the proportion of land impression recovered. It is
immediately obvious that the variability in feature expression is large
when only a small fraction of the land is recovered, such as 25\%. For
instance, \texttt{sum\_peaks} and \texttt{cms} both drop, while
\texttt{D} rises. Interestingly, some of the features are better
expressed for the middle-fixed case. Overall, feature expression remains
relatively consistent as long as we recover 50\% or more of the land
impression. Feature \ref{fig:featexp2} shows the feature expression for
known non-matches by comparison. The non-matches don't exhibit the same
pattern of better feature expression for the middle-fixed case, except
perhaps for very low degradation levels. However, feature expression
rises as a function of the land proportion, which indicates why the
random forest begins predicting more positives, raising the sensitivity,
but drastically lowering the specificity.

@@

\begin{figure}[htbp]
\centering
\includegraphics{degraded-paper_files/figure-latex/featexp2-1.pdf}
\caption{\label{fig:featexp2}Feature expression for known non-matches, as a
function of land proportion. As expected by the fact that the false
positive rate increases as the land proportion decreases, so too does
the feature expression. However, unlike for the known matches, fixing
the middle portion of the land does not seem to lead to more expressed
features, except perhaps for very low degradation levels.}
\end{figure}

To come full circle, we now attempt to match a particular land which
exhibits bad tank rash. Figure \ref{fig:br924} provides an image of the
surface of this land impression. Due to the tank rash, this particular
land impression was originally excluded from consideration (see (Hare,
Hofmann, and Carriquiry 2016)) based on our subjective assessment of the
quality of the land. However, it appears that approximately half of the
bullet land remains relatively unaffected. We extract a signature from
the unaffected half and attempt to match this signature to its full
known match.

@@

\begin{align*}
P(M) &= 1 - P(NM) \\
     &= 1 - (P(NM1) \times P(NM2) \times ... \times P(NM6)) \\
     &= 1 - ((1 - P(M1)) \times (1 - P(M2)) \times ... \times (1 - P(M6)))
\end{align*}
